Testing and Isolation Efficacy:\\Insights from a Simple Epidemic Model

### 2022-03-13 manuscript_BMB.tex
Bolker comments and tweaks

\institute{Ali Gharouni \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada \\
              Tel.: +100-00-000000\\
              Fax: +100-00-000000\\
              \email{agharoun@uottawa.ca}\\
              \emph{Present address:}
              Department of Mathematics and Statistics, University of Ottawa, Ottawa, Ontario \\%  if needed\\
              \email{agharoun@uottawa.ca}\\
           \and
           Fred .M. Abdelmalek \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada \\
              \email{abdelf1@mcmaster.ca}\\  
           \and
           David J. D. Earn \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada\\
              Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Canada\\
              \email{earn@math.mcmaster.ca}\\
           \and
          Jonathan Dushoff \at
              Department of Biology, McMaster University, Hamilton, Canada\\
              Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Canada\\
              South African Centre for Epidemiological Modelling and Analysis, University of Stellenbosch, Stellenbosch,               South Africa\\
              \email{dushoff@mcmaster.ca} \\
           \and
          Benjamin M. Bolker \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada \\
              Department of Biology, McMaster University, Hamilton, Canada\\
              Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Canada\\
              \email{bolker@mcmaster.ca}
}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; behavioural dynamics \citep{weitz2020awareness}); 
% reviewer2:com1
Further, our model ignores the fact that the testing and isolation parameters can change over time due to public health responses. In other words, our model makes an assumption of ``independence'': we assume that testing and isolation affect the epidemic progression, but we do not account for the fact that epidemic progression is also likely to affect testing and isolation strategies.%
Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

Many of the qualitative results we have derived confirm simple, common-sense intuitions. In particular, we can generally decrease $\Rnum$ by increasing isolation efficacy or testing intensity; returning tests faster, if individuals do not isolate while they are waiting for results; or increasing testing focus to target individuals who are likely to be infectious (e.g., symptomatic people or close contacts of known infections).

However, we did find two surprising patterns: under some conditions longer delays in returning tests can reduce epidemic spread, and increasing testing rates can increase spread.

Over broad regions of parameter space, decreasing $\omega$---i.e., \emph{slowing} the rate at which test results are returned---\emph{decreases} $\Rnum$ (for random testing, the parameter region is $\theta\_w \gtrsim 0.25$; for targeted testing, $\theta\_w \geq 0.25$ and either $\theta\_c \ge 0.5$ or $1/\omega > 5$; see Table~\ref{tab:params} for parameter definitions).  This result is counterintuitive and would not be expected by public health authorities that have invested a great deal of effort in reducing delays from testing to results. Dynamically, this effect occurs because speeding up test returns shortens the isolation period of uninfected individuals (for infected people it only shortens the time to progression to the isolation level of the confirmed-positive compartment). Slowing test returns increases $\Rnum$ only if the proportion of infectives in the tested population is high and isolation is relatively strong among people waiting for test results.

@@

% reviewer1:com2
Future steps should explore mathematically tractable ways to model some of these factors more precisely. For example, separating the infected class into exposed, symptomatic, and a- or pre-symptomatic compartments and allowing the testing weights to vary across non-symptomatic (exposed/asymptomatic/presymptomatic) vs.\ symptomatic compartments could reflect the allocation of tests for diagnostic purposes (targeting symptomatic individuals) vs.\ contact-tracing (targeting infected but non-symptomatic individuals) vs.\ screening (relatively equal weights, depending on the venue). This could also allow exploration of different assumptions about individuals' likelihood of testing positive in different compartments, with implications for the questions about repeat testing discussed above. Alternatively, one could make the testing weights depend on the testing intensity or test-return rate as suggested above. Whatever complexity is added would probably put the model beyond reach of the analytical methods we have used in this paper, but one could still use semi-numerical methods such as constructing the next-generation matrix and using it to evaluate the derivatives of $\Rnum$ with respect to the parameters numerically.

>>>>

\institute{Ali Gharouni \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada \\
              Tel.: +100-00-000000\\
              Fax: +100-00-000000\\
              \email{agharoun@uottawa.ca}\\
              \emph{Present address:}
              Department of Mathematics and Statistics, University of Ottawa, Ottawa, Ontario \\%  if needed\\
              \email{agharoun@uottawa.ca}\\
           \and
           Fred M. Abdelmalek \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada \\
              \email{abdelf1@mcmaster.ca}\\  
           \and
           David J. D. Earn \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada\\
              Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Canada\\
              \email{earn@math.mcmaster.ca}\\
           \and
          Jonathan Dushoff \at
              Department of Biology, McMaster University, Hamilton, Canada\\
              Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Canada\\
              South African Centre for Epidemiological Modelling and Analysis, University of Stellenbosch, Stellenbosch,               South Africa\\
              \email{dushoff@mcmaster.ca} \\
           \and
          Benjamin M. Bolker \at
              Department of Mathematics and Statistics, McMaster University, Hamilton, Canada \\
              Department of Biology, McMaster University, Hamilton, Canada\\
              Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Canada\\
              \email{bolker@mcmaster.ca}
}

@@

In this paper, we have developed and analyzed a simple compartmental model that combines epidemiological dynamics---as defined by a simple SIR model---with the dynamics of testing and isolation. Our model is a caricature: it models the most basic feedbacks between epidemic and testing processes, but does not attempt to incorporate the many known complications of COVID-19 epidemiology (e.g., exposed, pre-symptomatic, and asymptomatic compartments \citep{kain2021chopping}; time-varying testing rates; and behavioural dynamics \citep{weitz2020awareness}). 
% reviewer2:com1
Further, our model ignores the fact that the testing and isolation parameters can change over time due to public health responses. In other words, we assume dynamical influences only work in one direction: we recognize that testing and isolation affect epidemic dynamics, but we do not account for the fact that epidemic dynamics are also likely to affect testing and isolation strategies.%
Thus, it is most appropriate for assessing the \emph{qualitative} phenomena that arise from the interactions between transmission dynamics and testing, rather than for making quantitative predictions or guiding pandemic responses.

Many of the qualitative results we have derived confirm simple, common-sense intuitions. In particular, it is not surprising that we can generally decrease $\Rnum$ by increasing isolation efficacy or testing intensity; returning tests faster, if individuals do not isolate while they are waiting for results; or increasing testing focus to target individuals who are likely to be infectious (e.g., symptomatic people or close contacts of known infections).

However, we did find two surprising phenomena: under some conditions longer delays in returning tests can reduce epidemic spread, and increasing testing rates can increase spread.

Over broad regions of parameter space, decreasing $\omega$---i.e., \emph{slowing} the rate at which test results are returned---\emph{decreases} $\Rnum$ (for random testing, the parameter region is $\theta\_w \gtrsim 0.25$; for targeted testing, $\theta\_w \geq 0.25$ and either $\theta\_c \ge 0.5$ or $1/\omega > 5$; see Table~\ref{tab:params} for parameter definitions).  This result is counterintuitive and would not be expected by public health authorities who have invested a great deal of effort in reducing delays from testing to results. Dynamically, this effect occurs because speeding up test returns shortens the isolation period of uninfected individuals (for infected people it only shortens the time to progression to the isolation level of the confirmed-positive compartment). Slowing test returns increases $\Rnum$ only if the proportion of infectives in the tested population is high and isolation is relatively strong among people waiting for test results.

@@

% reviewer1:com2
Future steps should explore mathematically tractable ways to model some of these factors more precisely. For example, separating the infected class into exposed, symptomatic, and a- or pre-symptomatic compartments and allowing the testing weights to vary across non-symptomatic (exposed\slash asymptomatic\slash presymptomatic) vs.\ symptomatic compartments could reflect the allocation of tests for diagnostic purposes (targeting symptomatic individuals) vs.\ contact-tracing (targeting infected but non-symptomatic individuals) vs.\ screening (relatively equal weights, depending on the venue). This development would also allow exploration of different assumptions about individuals' likelihood of testing positive in different compartments, with implications for the questions about repeat testing discussed above. Alternatively, one could make the testing weights depend on the testing intensity or test-return rate as suggested above. Furthermore, one could allow for differential isolation, and different transmission probabilities, in different compartments. Whatever complexity is added would probably put the model beyond reach of the analytical methods we have used in this paper, but one could still use semi-numerical methods such as constructing the next-generation matrix and using it to compute the derivatives of $\Rnum$ with respect to the parameters for particular choices or ranges of parameter values.

### 2022-03-13 response1_BMB.tex
once through
added Muler to refs, marked %reviewer:com in the mans
Bolker comments and tweaks

\usepackage{amssymb,amsmath,amsthm}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}\renewcommand\thelinenumber{\color{gray}\arabic{linenumber}}

@@

\res 1 
% {see "Muller" in manuscript_BMB.tex}

@@

\res 2
[not sure here!]

% JD try to write text
% {search for reviewer1:com2 in the manuscript_BMB.tex}

{\it While I think that the analysis already presented in the paper is sufficient for the current paper, I think that ideas related to these questions might be interesting for future research or as points for discussion.}
\com 3 
{\it Due to its important effects on R0, I think it could be useful to describe the rate $\omega$ for return of test results directly in the text (in addition to the table of parameters) before it appears in Equation 4 in the expression for the disease-free equilibrium.}

@@

% [1] K. Muller and P. A. Muller, ``Mathematical modelling of the spread of covid-19 on a university campus,'' Infectious Disease Modelling, vol. 6, pp. 1025-1045, 2021.
% ~/Downloads/MullMull.pdf
% https://pubmed.ncbi.nlm.nih.gov/34414342/

@@

\res 1
We now have added a line in the manuscript (the first paragraph of the Discussion section) to clarify what we mean by the term ``caricature''. 
% {search for reviewer2:com2 in the manuscript_BMB.tex}

>>>>

\usepackage{url}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{xcolor}
\usepackage{xspace}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}\renewcommand\thelinenumber{\color{gray}\arabic{linenumber}}

@@

\res 1 We now talk more generally about cautious behavior of people waiting (and cite Muller and Muller 2021 on this topic) in our discussion. 
% {reviewer1:com1 in the manuscript_BMB.tex}

@@

\res 2 This is a good point. We now discuss possible interactions between epidemiological status (as reflected in a more complex model) and testing in the Discussion as a possible future investigation.
% JD try to write text
% {search for reviewer1:com2 in the manuscript_BMB.tex}

{\it While I think that the analysis already presented in the paper is sufficient for the current paper, I think that ideas related to these questions might be interesting for future research or as points for discussion.}

\com 3 
{\it Due to its important effects on R0, I think it could be useful to describe the rate $\omega$ for return of test results directly in the text (in addition to the table of parameters) before it appears in Equation 4 in the expression for the disease-free equilibrium.}

@@

[1] K. Muller and P. A. Muller, ``Mathematical modelling of the spread of COVID-19 on a university campus,'' Infectious Disease Modelling, vol. 6, pp. 1025-1045, 2021.
 \url{https://pubmed.ncbi.nlm.nih.gov/34414342/}

@@

\res 1
We now have added a line in the manuscript (the first paragraph of the Discussion section) to clarify what we mean by the term ``caricature''. 
% {search for reviewer2:com1 in the manuscript_BMB.tex}
